---
title: "4: Instrumental Variables"
subtitle: "Quantitative Causal Inference"
author: "<large>J. Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "April 23 and 28, 2026"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes the bibliography
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
#myBib <- ReadBib("assets/myBib.bib", check = FALSE)
# Note: don't forget to clear the knitr cache to account for changes in the
# bibliography.

peruemotions <- read.csv("https://github.com/jnseawright/PS406/raw/main/data/peruemotions.csv")
```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

###Today's Plan

1.  Why Instrumental Variables (IV)?
2.  The IV estimator and its assumptions
3.  Two-stage least squares
4.  Diagnostics
5.  Heterogeneity and LATE interpretation
6.  Applications and limitations

---
### Endogeneity in OLS

-   $E(\mathbf{u} | \mathbf{X}) = 0$?

-   As you'll recall,
    $E(\hat{\mathbf{\beta}}) = \beta + (\mathbf{X}^{T} \mathbf{X})^{-1} E(\mathbf{X}^{T} \mathbf{u})$.
    So, if $E(\mathbf{X}^{T} \mathbf{u}) = \mathbf{\nu} \neq 0$, then
    $E(\hat{\mathbf{\beta}} - \mathbf{\beta}) = (\mathbf{X}^{T} \mathbf{X})^{-1} \mathbf{\nu} \neq 0$.

---
### Consequences of Endogeneity

-   When $\mathbf{X}$ is endogenous, then our estimates of
    $\hat{\mathbf{\beta}}$ will be a mixture of the desired relationship
    between $\mathbf{X}$ and $\mathbf{y}$ *and* the nuisance
    relationship between $\mathbf{X}$ and $\mathbf{u}$.

---
### How Can Endogeneity Arise?

-   Omitted explanatory variables
-   Measurement error on the right-hand side of the model
-   Simultaneity between the right- and left-hand sides of the model
-   etc.

---
### What to Do When Endogeneity Is a Problem?

1.  Give up.

2.  Try to change the model by including all omitted relevant variables.

3.  Find an instrument.

4.  Find other data.

---
### Instrumental Variables

-   Suppose the model is:
    $\mathbf{y} = \mathbf{W} \mathbf{\gamma} + \mathbf{x} \beta + \mathbf{\epsilon}$.
    The $\mathbf{W}$ variables are exogenous, but the $\mathbf{x}$
    variable is endogenous.

-   Now, assume that there exists a variable $z$ that *doesn't* belong
    in the regression model, with the following two characteristics:

    -   $cov(\mathbf{z}^{T} \mathbf{x}) \neq 0$

    -   $E(\mathbf{z}^{T} \mathbf{\epsilon}) = 0$

---
### Instrumental Variables

If these conditions are met (doesn't belong in the regression, related
linearly with $\mathbf{x}$, no connection with $\mathbf{\epsilon})$,
then $\mathbf{z}$ meets the mathematical definition of an *instrument*.

---
###Three Core Assumptions

1.  Relevance: $\mathrm{Cov}(Z,X) \neq 0$
2.  Exclusion restriction: $Z$ affects $Y$ only through $X$
3.  Independence / exogeneity: $Z$ is as good as randomly assigned (or at least uncorrelated with unobservables)

---
### Relevance Failure

```{r, echo = FALSE, out.width="70%", fig.retina = 1}
library(dagitty)
relevancefailure.dag <- dagitty('dag {
    Instrument [pos="0,1"]
    Treatment [pos="1,2"]
    Outcome [pos="2,2"]
    Confounder_1 [pos="1,0"]
    Confounder_2 [pos="2,1"]
    
    Confounder_1 -> Treatment <- Confounder_2 
    Confounder_1 -> Outcome <- Confounder_2 
    Treatment -> Outcome  
}')
plot(relevancefailure.dag)
```

---
### Exclusion Restriction Failure
```{r, echo = FALSE, out.width="70%", fig.retina = 1}
library(dagitty)
exclusionfailure.dag <- dagitty('dag {
    Instrument [pos="0,1"]
    Treatment [pos="1,2"]
    Outcome [pos="2,2"]
    Confounder_1 [pos="1,0"]
    Confounder_2 [pos="2,1"]
    
    Confounder_1 -> Treatment <- Confounder_2 
    Confounder_1 -> Outcome <- Confounder_2 
    Instrument -> Treatment -> Outcome  
    Instrument -> Outcome  
}')
plot(exclusionfailure.dag)
```

---
### Exogeneity Failure
```{r, echo = FALSE, out.width="70%", fig.retina = 1}
library(dagitty)
exogeneityfailure.dag <- dagitty('dag {
    Instrument [pos="0,1"]
    Treatment [pos="1,2"]
    Outcome [pos="2,2"]
    Confounder_1 [pos="1,0"]
    Confounder_2 [pos="2,1"]
    
    Confounder_1 -> Treatment <- Confounder_2 
    Confounder_1 -> Outcome <- Confounder_2 
    Instrument -> Treatment -> Outcome  
    Confounder_1 -> Instrument
}')
plot(exogeneityfailure.dag)
```

---
### DAGs and Finding Instruments

If we can specify our causal structure well in advance, mathematical graph theory can help identify which variables are instruments.
---
```{r, echo = TRUE, out.width="70%", fig.retina = 1}
library(dagitty)
hypotheticalinstruments.dag <- dagitty( "dag { Polarization -> DemocraticErosion ElitePower -> DemocraticErosion Corruption -> Polarization Corruption -> DemocraticErosion SocialMedia -> Polarization PrimaryElections -> Polarization PrimaryElections -> ElitePower EconomicInequality -> ElitePower EconomicInequality -> Polarization Polarization [exposure]
DemocraticErosion [outcome]}" )
```

---
```{r, echo = TRUE, out.width="50%", fig.retina = 1}
plot( hypotheticalinstruments.dag )
```
---
```{r, echo = TRUE, out.width="50%", fig.retina = 1}
instrumentalVariables(hypotheticalinstruments.dag)
```

---
### Bivariate IV

-   Let's momentarily consider a bivariate regression,
    $\mathbf{y} = \mathbf{x} \beta + \mathbf{\epsilon}$, with instrument
    $\mathbf{z}$.

-   The OLS estimate of $\beta$ is
    $(\mathbf{x}^{T}\mathbf{x})^{-1} \mathbf{x}^{T}\mathbf{y}$.

-   Consider instead the IV estimate of $\beta$:
    $(\mathbf{z}^{T}\mathbf{x})^{-1} \mathbf{z}^{T}\mathbf{y}$.

-   $E(\hat{\beta}_{IV}) = E((\mathbf{z}^{T}\mathbf{x})^{-1} \mathbf{z}^{T}\mathbf{y}) = E((\mathbf{z}^{T}\mathbf{x})^{-1} \mathbf{z}^{T} [\mathbf{x} \beta + \mathbf{\epsilon}]) $

-   $E(\hat{\beta}_{IV}) = E((\mathbf{z}^{T}\mathbf{x})^{-1} \mathbf{z}^{T} \mathbf{x} \beta) + E((\mathbf{z}^{T}\mathbf{x})^{-1} \mathbf{z}^{T} \mathbf{\epsilon}) = \beta + 0$


---
### Instrumental Variables

-   Now let's consider a multivariate regression,
    $\mathbf{Y} = \mathbf{X} \mathbf{\beta} + \mathbf{\epsilon}$, with
    some $t \leq k$ of the $\mathbf{X}$ variables endogenous, and with
    $t$ instruments $\mathbf{z}_{1} \ldots \mathbf{z}_{t}$.

---
### Multivariate IV

-   The OLS estimate of $\mathbf{\beta}$ is
    $(\mathbf{X}^{T}\mathbf{X})^{-1} \mathbf{X}^{T}\mathbf{y}$.

-   Form the matrix $\mathbf{Z}$, containing the $t$ instruments, as
    well as the $k - t$ exogenous elements from $\mathbf{X}$.

-   The IV estimate of $\mathbf{\beta}$ is:
    $(\mathbf{Z}^{T}\mathbf{X})^{-1} \mathbf{Z}^{T}\mathbf{y}$.

---
### Multivariate IV

-   As in the bivariate situation, given the IV assumptions, the IV
    estimator eliminates the problem of endogeneity.

-   This estimator only works if the number of instruments is exactly
    equal to the number of endogenous variables.

---
### A Brief History of Instrumental Variables

-   The method of instrumental variables has surprisingly deep roots in econometrics.

-   **Philip Wright (1928)** is credited with the first explicit use of IV in his book *The Tariff on Animal and Vegetable Oils*.

-   Wright faced a classic simultaneity problem: estimating supply and demand curves for flaxseed oil.

    -   Price and quantity are jointly determined.
    -   OLS would give a mixture of supply and demand elasticities.

-   His solution: use exogenous shifters of supply (e.g., weather) to trace out the demand curve, and shifters of demand (e.g., tariff changes) to trace out the supply curve.

---
### Wright's Insight

-   Wright realized that if you have a variable that shifts one curve but not the other, you can identify the other curve's parameters.

-   This is exactly the modern IV intuition: an instrument $Z$ affects $X$ (e.g., quantity supplied) but has no direct effect on $Y$ (e.g., price) except through $X$.

-   Wright's work remained largely unknown for decades; it was rediscovered by econometricians in the 1970s and 1980s.

-   Today, IV is one of the most widely used methods for causal inference with observational data.

---
### Wright's Flaxseed Oil Example: A Simple Simulation

```{r wright_sim, echo = TRUE, fig.height = 4}
# Simulate a supply-demand system with an instrument
set.seed(2026)
n <- 1000

# Instrument: weather shock (shifts supply)
weather <- rnorm(n)

# Supply shifter (weather) and demand shifter (income) 
# but we only observe weather as instrument
income <- rnorm(n)  # unobserved confounder?

# True demand: P = 10 - 0.5*Q + income*0.3 + epsilon_d
# True supply: P = 2 + 0.8*Q - weather*0.4 + epsilon_s

# Solve for equilibrium Q and P (simplified)
Q <- 5 + 0.2*income - 0.3*weather + rnorm(n)
P <- 8 - 0.4*Q + 0.3*income + rnorm(n)

# OLS of P on Q (tries to estimate demand, but biased)
summary(lm(P ~ Q))

# IV using weather as instrument for Q
library(ivreg)
summary(ivreg(P ~ Q | weather))
```

-   OLS gives a biased estimate of the demand elasticity (price-quantity relationship).
-   IV using weather (a supply shifter) recovers the demand curve.


---
### Intuition: Two-Stage Least Squares

IV can be thought of as a two-step process:

1. **First stage**: Regress the endogenous X on the instrument Z (and controls). This isolates the exogenous variation in X—the part predicted by Z.
2. **Second stage**: Regress Y on the predicted values from the first stage.

This "purified" X is no longer correlated with the error term, giving us consistent estimates.

The algebra on the next slide shows why this is equivalent to the IV estimator we already derived.

---
### 2SLS

-   Let's partition the independent variables into two matrices,
    $\mathbf{W}$, which has the $k - t$ exogenous variables in the model
    of $\mathbf{y}$, and $\mathbf{X}$, which has the $t$ endogenous
    variables.

-   So the $\mathbf{Z}$ matrix is the $\mathbf{W}$ matrix with $t$ extra
    columns containing the instruments.

---
### 2SLS

-   Suppose we regress each column of the $\mathbf{X}$ matrix on the
    matrix $\mathbf{Z}$ and form the fitted values.

-   $\hat{\mathbf{X}} = \mathbf{Z} (\mathbf{Z}^{T} \mathbf{Z})^{-1} \mathbf{Z}^{T} \mathbf{X}$

-   Now use $\hat{\mathbf{X}}$ in the place of $\mathbf{X}$ in the OLS
    regression formula.

---
### 2SLS

$$
\begin{split}
\hat{\mathbf{\beta}}_{IV} = & (\mathbf{X}^{T}\mathbf{Z} (\mathbf{Z}^{T} \mathbf{Z})^{-1} \mathbf{Z}^{T} \mathbf{Z} (\mathbf{Z}^{T} \mathbf{Z})^{-1} \mathbf{Z}^{T} \mathbf{X})^{-1} \\
& \mathbf{X}^{T}\mathbf{Z} (\mathbf{Z}^{T} \mathbf{Z})^{-1} \mathbf{Z}^{T} \mathbf{y} = \\
& (\mathbf{X}^{T}\mathbf{Z} (\mathbf{Z}^{T} \mathbf{Z})^{-1} \mathbf{Z}^{T} \mathbf{X})^{-1} \\ & \mathbf{X}^{T}\mathbf{Z} (\mathbf{Z}^{T} \mathbf{Z})^{-1} \mathbf{Z}^{T} \mathbf{y} = \\
& (\mathbf{Z}^{T}\mathbf{X})^{-1} \mathbf{Z}^{T}\mathbf{y}
\end{split}
$$

-   The instrumental variables estimator gives the same coefficient
    estimates as running an OLS regression using $\hat{\mathbf{X}}$ as predicted by $\mathbf{Z}$ in
    the place of $\mathbf{X}$.

---
### Variance in Instrumental Variables

-   $\hat{\mathbf{X}}$ is a random variable, so the normal OLS standard
    errors will underestimate uncertainty when using IV.

-   Instead, the correct estimate of the standard errors of the
    coefficient estimates in IV is:

    -   $\hat{V} (\hat{\mathbf{\beta}}_{IV}) = \hat{\sigma}^{2} (\mathbf{Z}^{T} \mathbf{X})^{-1} \mathbf{Z}^{T} \mathbf{Z} (\mathbf{X}^{T} \mathbf{Z})^{-1}$

---
### Examples of Proposed Instruments

-   Suppose we're interested in the relationship between education and
    some political variable.

    -   One proposed instrument for education, due to David Card (1995),
        is residential proximity to a college or university.

    -   A second proposed instrument for education, due to Angrist and
        Krueger (1991) is month of birth.

    -   A third instrument, from Nguyen et al. (2016), involves genetic
        risk score for years of schooling.

---
### Examples of Proposed Instruments

-   Suppose our focus is on the relationship between economic
    performance and civil war in agricultural countries.

    -   Miguel, Satyanath, Sergenti, E. (2004) suggest using rainfall as
        an instrument for economic performance.

---
```{r, echo = TRUE, out.width="40%", fig.retina = 1, fig.align='center'}
library(haven)
mss_repdata_1_ <- read_dta("https://github.com/jnseawright/PS406/raw/main/data/mss_repdata%20(1).dta")
```

---
```{r, echo = TRUE, out.width="40%", fig.retina = 1, fig.align='center'}
library(ivreg)
migueliv <- ivreg(any_prio ~ gdp_g + gdp_g_l + y_0 + polity2l + ethfrac + relfrac + Oil + lpopl1 + lmtnest | GPCP_g + GPCP_g_l+ y_0 + polity2l + ethfrac + relfrac + Oil + lpopl1 + lmtnest, data=mss_repdata_1_)
summary(migueliv)
```

---
### Reading IV Regression Output

Key elements to examine:

1. **Coefficient on endogenous variable**: `gdp_g` = 0.064 — a 1 percentage point increase in growth increases conflict probability by 6.4 points? (Check units—growth is in percent, conflict is binary)

2. **Standard errors**: Clustered SEs are larger (0.038 vs. 0.027) — accounting for within-country correlation matters

3. **First-stage statistics** (not shown here but available via `summary(..., diagnostics=TRUE)`): Check instrument strength

4. **Model fit**: IV $R^2$ can be negative—don't interpret as usual

---
```{r, echo = TRUE, out.width="40%", fig.retina = 1, fig.align='center'}
library(lmtest)
library(sandwich)
```

---
```{r, echo = TRUE, out.width="40%", fig.retina = 1, fig.align='center'}
coeftest(migueliv, vcov = vcovCL(migueliv, cluster = ~country_name))
```

---
### What's Wrong with Weak Instruments?

-   For an IV estimate of a regression with only one independent
    variable and only one instrument, the IV estimator is:
    $(\mathbf{z}^{T} \mathbf{x})^{-1} \mathbf{z}^{T} \mathbf{y}$, which
    is the same as
    $cov(\mathbf{z}, \mathbf{y})/cov(\mathbf{z}, \mathbf{x})$.

---
### What's Wrong with Weak Instruments?

-   The $cov(\mathbf{z}, \mathbf{y})$ may be thought of as a combination
    of three components:

    -   the direct effect of $\mathbf{z}$ on $\mathbf{y}$,

    -   the indirect effect of $\mathbf{z}$ on $\mathbf{y}$ via
        $\mathbf{x}$,

    -   and any correlation between $\mathbf{z}$ and $\mathbf{u}$.

---
### What's Wrong with Weak Instruments?

-   If $cov(\mathbf{z}, \mathbf{x})$ is big, then a moderate amount of
    contamination of $cov(\mathbf{z}, \mathbf{y})$ with undesirable
    information will have only a small effect on the estimate.

-   If $cov(\mathbf{z}, \mathbf{x})$ is very small, then even a small
    amount of contamination of $cov(\mathbf{z}, \mathbf{y})$ with
    undesirable information will lead to serious bias in the estimate.

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
summary(migueliv)
```

---
### What if the exclusion restriction fails?

The bias formula for instrumental variables is:

$$\mathrm{plim} \hat{\beta}_{IV} = \beta + \frac{\mathrm{Cov}(Z,u)/\mathrm{Var}(Z)}{\mathrm{Cov}(Z,X)/\mathrm{Var}(Z)}$$
---
### What if the exclusion restriction fails?

- Even a small $\mathrm{Cov}(Z,u)$ can cause large bias if $\mathrm{Cov}(Z,X)$ is small (weak instruments). 

- If $\mathrm{Cov}(Z,X)$ is large but $\mathrm{Cov}(Z,u)$ is also large (invalid instrument) we can also end up with large bias.

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
miguellm2 <- lm(any_prio ~ gdp_g + gdp_g_l +GPCP_g + GPCP_g_l + y_0 + polity2l + ethfrac + relfrac + Oil + lpopl1 + lmtnest + year:country_name, data=mss_repdata_1_)

summary(miguellm2)
```

---
### Diagnostics in Instrumental Variables

-   After estimating an IV model, we must diagnose whether the instrument(s) are valid and strong enough.

-   Four key diagnostic families:

    1.  **First‑stage diagnostics**: relevance of instruments.

    2.  **Overidentification tests**: validity when instruments > endogenous variables.

    3.  **Weak instrument robust tests**: inference that remains valid with weak instruments.

    4.  **Endogeneity tests**: whether IV is actually needed.

-   Each addresses a different threat to IV consistency.

---

### First‑Stage Diagnostics: Relevance

-   The first stage is the regression of the endogenous variable(s) on all instruments (and exogenous covariates).

-   For a single endogenous regressor, the **first‑stage F‑statistic** tests whether the instruments jointly have explanatory power.

-   A common rule of thumb (Stock & Yogo, 2005): F < 10 indicates weak instruments → IV bias can be large.

-   Also useful: **partial R²** – the share of variation in the endogenous variable explained by the instruments after controlling for exogenous variables.

---
```{r firststage, echo = TRUE, warning = FALSE, message = FALSE}
# First-stage regression for the Miguel et al. example
# Endogenous: gdp_g (growth), instrument: GPCP_g (rainfall)
# Include all exogenous controls as in the original model

first_stage <- lm(gdp_g ~ GPCP_g + GPCP_g_l + y_0 + polity2l + ethfrac + relfrac + Oil + lpopl1 + lmtnest, 
                  data = mss_repdata_1_)

# F-statistic for the instrument (GPCP_g) – use linearHypothesis from car package
library(car)
f_test <- linearHypothesis(first_stage, c("GPCP_g = 0"))
f_test
```

---
```{r firststage_partialR2, echo = TRUE}
# Partial R²: proportion of variance explained by instruments after controls
# Full model R²
r2_full <- summary(first_stage)$r.squared

# Model without the instrument
first_stage_noZ <- lm(gdp_g ~ GPCP_g_l + y_0 + polity2l + ethfrac + relfrac + Oil + lpopl1 + lmtnest, 
                      data = mss_repdata_1_)
r2_noZ <- summary(first_stage_noZ)$r.squared

# Partial R²
partial_r2 <- r2_full - r2_noZ
partial_r2
```

-   The F‑statistic on the instrument is `r round(f_test[2, "F"], 2)` with a p‑value of `r round(f_test[2, "Pr(>F)"], 4)`.

-   This F is well above 10, suggesting the instrument is reasonably strong. (But recall: strength alone does not guarantee validity.)

---

### Overidentification Tests

-   When the number of instruments ($m$) exceeds the number of endogenous variables ($k$), we have **overidentification**.

-   The extra instruments allow us to test the joint validity of all instruments (the exclusion restriction) – under the assumption that at least $k$ instruments are valid.

-   Common tests:
    -   **Sargan test** (homoskedastic errors)
    -   **Hansen J test** (heteroskedasticity‑robust)

-   Null hypothesis: all instruments are valid (i.e., uncorrelated with the error term).

-   Rejection implies at least one instrument is invalid, but the test cannot tell us which one.

---
```{r overid, echo = TRUE}
# The ivreg package can produce Sargan (or Hansen) test with summary()
summary(migueliv, diagnostics = TRUE)
```

-   The diagnostic output includes:
    -   **Sargan test**: statistic `r round(summary(migueliv, diagnostics = TRUE)$diagnostics["Sargan", "statistic"], 3)`, p‑value `r round(summary(migueliv, diagnostics = TRUE)$diagnostics["Sargan", "p-value"], 3)`.

-   With p > 0.05, we do **not** reject the null. This is consistent with instrument validity, but we must remember the test’s limitations: low power, and it assumes at least one instrument is valid.

---

### Weak Instrument Robust Tests

-   If instruments are weak, conventional IV standard errors and confidence intervals can be misleading – coverage rates can be far from nominal.

-   **Weak‑instrument‑robust tests** provide inference that remains valid regardless of instrument strength.

-   Popular choices:
    -   **Anderson–Rubin (AR) test**: tests the structural parameter $\beta$ by examining the reduced form. It is robust to weak instruments but can have low power when many instruments are weak.
    -   **Conditional Likelihood Ratio (CLR) test** (Moreira, 2003): often more powerful than AR.

-   In R, the `ivreg` package does not implement these directly, but the `AER::ivreg` function (same as `ivreg`) can be used with the `diagnostics = TRUE` option, which includes a weak‑instrument test (Cragg–Donald F) and an AR confidence interval if requested.

---
```{r weakrobust, echo = TRUE}
# For AR confidence intervals, we can use the ivreg package with the `AR` option
# This requires a model with one endogenous variable and possibly multiple instruments
library(ivreg)
ar_ci <- confint(migueliv, type = "AR")
ar_ci
```

-   The Anderson–Rubin 95% confidence interval for `gdp_g` is `[`r round(ar_ci[1,1], 3)`, `r round(ar_ci[1,2], 3)`]`. This interval is valid even if instruments are weak.

-   Compare with the conventional IV confidence interval: `r round(confint(migueliv)[2,1], 3)` to `r round(confint(migueliv)[2,2], 3)`. The AR interval is slightly wider, reflecting the additional uncertainty under weak‑instrument asymptotics.

---

### Testing for Endogeneity

-   If a variable is actually exogenous, OLS is more efficient than IV. Testing for endogeneity helps decide whether IV is necessary.

-   **Hausman test** (original): compares OLS and IV estimates. Under the null of exogeneity, both are consistent but OLS is efficient; under the alternative, only IV is consistent. Large differences suggest endogeneity.

-   **Durbin–Wu–Hausman (DWH) test** (regression‑based form): easier to implement and robust to heteroskedasticity.

-   Important caveat: the test requires that the instruments are valid. If instruments are invalid, the test can be misleading.

---
```{r endogeneity, echo = TRUE}
# Regression-based Durbin-Wu-Hausman test
# Step 1: First stage residuals
first_stage_resid <- resid(first_stage)

# Step 2: Include residuals in the structural model
dwh_model <- lm(any_prio ~ gdp_g + gdp_g_l + y_0 + polity2l + ethfrac + relfrac + Oil + lpopl1 + lmtnest + first_stage_resid, 
                data = mss_repdata_1_)

# Test significance of first_stage_resid
summary(dwh_model)$coefficients["first_stage_resid", ]
```

-   The coefficient on the first‑stage residual is `r round(summary(dwh_model)$coefficients["first_stage_resid", "Estimate"], 4)` with p = `r round(summary(dwh_model)$coefficients["first_stage_resid", "Pr(>|t|)"], 4)`. A significant residual indicates that OLS would be inconsistent – i.e., `gdp_g` is endogenous.

-   Here, the residual is not significant (p > 0.05), suggesting that OLS might actually be acceptable for this variable. But this test hinges on instrument validity.

---
### Interpreting the Endogeneity Test

The DWH test fails to reject the null that `gdp_g` is exogenous (p = 0.34).

**This does NOT mean**: 
- The instrument is valid (that's a separate assumption)
- IV is unnecessary (the test has low power)

**This DOES mean**:
- OLS and IV estimates are not statistically distinguishable
- If you believe the instrument is valid, OLS might be preferred for efficiency
- But the reduced form results suggest caution

---


### Summary: IV Diagnostics Checklist

| Diagnostic | Purpose | R implementation | Caution |
|------------|---------|------------------|---------|
| First‑stage F | Check instrument strength | `linearHypothesis()` in first‑stage lm | F > 10 is only a rule of thumb |
| Overidentification (Sargan/Hansen) | Test instrument validity | `summary(ivreg, diagnostics = TRUE)` | Assumes at least one instrument valid; low power |
| Weak‑instrument robust CI | Valid inference under weak IV | `confint(..., type = "AR")` | May be wide; choose test based on power |
| Endogeneity test (DWH) | Check if IV is needed | Regression of residuals | Requires valid instruments |

-   Always report first‑stage F and, if overidentified, an overidentification test.
-   Consider weak‑instrument robust inference if F is low.
-   Interpret all tests with caution: they are diagnostic tools, not definitive proof.
```



---
### Encouragement Designs in Experiments

-   Intent-to-treat analysis

-   Use the treatment assignment as an instrument, the actual treatment
    received as the treatment variable, and the outcome as normal.

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
library(readr)
peruemotions <- read_csv("https://github.com/jnseawright/PS406/raw/main/data/peruemotions.csv")
```

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
summary(lm(outsidervote~simpletreat, data=peruemotions))
```

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
summary(lm(outsidervote~enojado, data=peruemotions))
```

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
summary(ivreg(outsidervote~enojado|simpletreat,data=peruemotions))
```

---
```{r, echo = TRUE, fig.retina = 1, fig.align='center'}
summary(lm(outsidervote~enojado+simpletreat, data=peruemotions))
```


---
### LATE: What Does IV Actually Estimate?

-   In an encouragement design (e.g., randomized treatment assignment with imperfect compliance), the instrument $Z$ is random assignment, but the treatment $X$ is what subjects actually receive.

-   IV does **not** estimate the average treatment effect (ATE) for the whole population—unless certain strong assumptions hold.

-   Instead, under reasonable assumptions, IV estimates the **Local Average Treatment Effect (LATE)**—the effect for a specific subgroup: **compliers**.

---
### Four Compliance Types

-   Imagine a binary treatment $X$ and a binary instrument $Z$ (e.g., encouragement).

-   Each individual falls into one of four latent groups based on how they respond to the instrument:

| Type           | Behavior                                                   |
|----------------|------------------------------------------------------------|
| **Compliers**  | $X=1$ when $Z=1$, $X=0$ when $Z=0$ (do as encouraged)     |
| **Always‑takers** | $X=1$ regardless of $Z$                                |
| **Never‑takers**  | $X=0$ regardless of $Z$                                |
| **Defiers**    | $X=0$ when $Z=1$, $X=1$ when $Z=0$ (do opposite)          |

-   In most encouragement designs, we cannot observe an individual's type directly—we only see one potential treatment status.

---
### Key Assumptions for LATE

1.  **Independence**: $Z$ is as good as randomly assigned (unconfoundedness).
2.  **Exclusion**: $Z$ affects the outcome $Y$ only through the treatment $X$ (no direct effect).
3.  **First stage**: $Z$ has a nonzero average effect on $X$ (relevance).
4.  **Monotonicity**: No defiers exist (or the proportion of defiers is zero).  
    *In a one‑sided encouragement design (e.g., only encouragement can increase treatment), monotonicity is automatically satisfied.*

Under these assumptions, the IV estimand (Wald estimator) equals the **average treatment effect for compliers** (LATE).

---
### The Wald Estimator = ITT / Compliance Rate

-   The **Intention‑to‑Treat (ITT)** effect: the effect of being assigned to encouragement on the outcome:
    $$ITT = E[Y | Z=1] - E[Y | Z=0]$$

-   The **Compliance rate** (first stage): the effect of encouragement on treatment uptake:
    $$FS = E[X | Z=1] - E[X | Z=0]$$

-   The Wald (IV) estimator is the ratio:
    $$\hat{\beta}_{IV} = \frac{ITT}{FS}$$

-   Intuition: The ITT is diluted by non‑compliers. Scaling by the compliance rate recovers the effect for those who actually comply with encouragement.

---
### Why LATE? A Decomposition

-   Under monotonicity and exclusion, the ITT can be written as:
    $$ITT = \text{(Effect on compliers)} \times \text{(Proportion compliers)}$$
    because always‑takers and never‑takers are unaffected by $Z$ (exclusion) and defiers are absent.

-   Hence:
    $$\text{LATE} = \frac{ITT}{\text{Proportion compliers}} = \frac{ITT}{FS}$$

-   This shows that IV isolates the treatment effect **only for compliers**—not for always‑takers or never‑takers.

---
### Limitations of LATE

-   The LATE may **not generalize** to the entire population:
    -   Always‑takers might have different treatment effects than compliers.
    -   Never‑takers might also differ.

-   The instrument defines the subpopulation of compliers. Different instruments can identify different LATEs.

-   External validity: we cannot automatically extrapolate the LATE to other settings or populations without additional assumptions.

-   Policy relevance: if a policy targets the same compliers identified by the instrument, LATE is directly useful. Otherwise, caution is needed.

---
### Applying LATE: The Peru Emotions Experiment

Let's work through a concrete example to see how LATE works in practice.

We have:
- **Random assignment** to a treatment that primes anger (`simpletreat`)
- **Actual anger** (`enojado`) is only partially caused by the treatment
- **Outcome**: voting for an outsider candidate (`outsidervote`)

We'll compute each component step by step.

---
### Illustration: Peru Emotions Experiment

-   Recall the Peru emotions example:
    -   **Instrument $Z$**: `simpletreat` (randomized encouragement)
    -   **Treatment $X$**: `enojado` (anger emotion, possibly endogenous)
    -   **Outcome $Y$**: `outsidervote` (voting for an outsider candidate)

-   We suspect that `enojado` is endogenous (unobserved confounders like personality traits). The random assignment `simpletreat` is a potential instrument.

---
```{r peru_late_setup, echo = TRUE}
# Load data if not already done
# peruemotions <- read_csv("https://github.com/jnseawright/PS406/raw/main/data/peruemotions.csv")

# Keep only complete cases for simplicity (as in earlier slide)
# Note: We drop missing values for simplicity. In practice, consider multiple imputation or other methods.
peruemotionstrim <- na.omit(data.frame(
  enojado = peruemotions$enojado,
  outsidervote = peruemotions$outsidervote,
  simpletreat = peruemotions$simpletreat,
  Cuzco = peruemotions$Cuzco,
  age = peruemotions$age
))
```



---
### Step 1: Compute the First Stage (Compliance Rate)

```{r peru_firststage, echo = TRUE}
# First stage: effect of instrument on treatment
fs <- lm(enojado ~ simpletreat + Cuzco + age, data = peruemotionstrim)
summary(fs)

# Extract the coefficient on simpletreat (compliance rate)
compliance_rate <- coef(fs)["simpletreat"]
compliance_rate
```

-   The estimated compliance rate is `r round(compliance_rate, 3)`.  
    Being in the treatment group increases the probability of feeling anger by about 8.6 percentage points.

---
### Step 2: Compute the Intention‑to‑Treat (Reduced Form)

```{r peru_itt, echo = TRUE}
# Reduced form: effect of instrument on outcome
itt <- lm(outsidervote ~ simpletreat + Cuzco + age, data = peruemotionstrim)
summary(itt)

# ITT coefficient
itt_effect <- coef(itt)["simpletreat"]
itt_effect
```

-   The ITT effect is `r round(itt_effect, 3)`.  
    Being assigned to the treatment group increases outsider voting by about 0.6 percentage points (not statistically significant? we'll check p-value later).

---
### Step 3: Compute the Wald (IV) Estimate Manually

```{r peru_wald, echo = TRUE}
# Wald estimator = ITT / compliance rate
wald_iv <- itt_effect / compliance_rate
wald_iv
```

-   The manual IV estimate is `r round(wald_iv, 3)`.  
    This suggests that for compliers, feeling anger increases the probability of outsider voting by about 7.1 percentage points.

---
### Step 4: Compare with `ivreg` Output

```{r peru_ivreg_compare, echo = TRUE}
# Run ivreg
library(ivreg)
iv_fit <- ivreg(outsidervote ~ enojado + Cuzco + age | simpletreat + Cuzco + age, 
                data = peruemotionstrim)
summary(iv_fit)

# Extract IV coefficient for enojado
iv_coef <- coef(iv_fit)["enojado"]
iv_coef
```

-   The `ivreg` coefficient on `enojado` is `r round(iv_coef, 3)`, matching the manual Wald estimate.  
    (Small differences may arise due to covariate adjustment—here we included Cuzco and age in both stages, which is correct.)

---
### Step 5: Interpret as LATE

-   Under monotonicity (no defiers) and exclusion (simpletreat affects voting only through anger), the IV estimate of 0.071 is the **average treatment effect of anger on outsider voting for compliers**.

-   Who are the compliers?  
    People who become angry **only if** they are in the treatment group (i.e., their anger is triggered by the experimental prime). They are not always angry, nor never angry—they respond to encouragement.

-   This effect may differ from the effect on always‑takers (who would be angry regardless) or never‑takers (who never become angry). Thus, we cannot automatically generalize to the whole population.

---
### Summary: LATE in a Nutshell

-   IV with a binary instrument and binary treatment, under monotonicity, identifies the **LATE for compliers**.

-   The Wald estimator is the ratio of ITT to first stage.

-   Always report:
    -   First‑stage strength (F‑statistic, compliance rate)
    -   Interpretation of the complier group
    -   Caveats about external validity

-   In the Peru example, the complier group is those whose anger is activated by the treatment prime. The estimated effect is about 7 percentage points.

```

---
### Beyond LATE: Estimating the ATE

-   LATE is useful but limited: it only gives the treatment effect for compliers.

-   Often we want the **Average Treatment Effect (ATE)** for the entire population.

-   Aronow and Carnegie (2013) show that under certain assumptions, we can recover the ATE by reweighting using the **compliance score**.

-   This approach is implemented in the `icsw` package (Inverse Compliance Score Weighting).

---
### What Is the Compliance Score?

-   The **compliance score** $\pi_i$ is the probability that unit $i$'s treatment status would be higher under encouragement than under control:

    $$\pi_i = P(X_i(1) > X_i(0) | \mathbf{W}_i)$$

    where $X_i(1)$ is potential treatment when $Z=1$, $X_i(0)$ when $Z=0$, and $\mathbf{W}_i$ are covariates.

-   For binary treatment and binary instrument, the compliance score is simply the probability of being a **complier**, conditional on covariates.

-   Intuition: Units with high compliance scores are likely compliers; those with low scores are likely always‑takers or never‑takers.

---
### How Does Reweighting Estimate the ATE?

-   Aronow and Carnegie show that if we weight observations by the inverse of the compliance score (for the treated) and by the inverse of $(1 - \pi_i)$ (for the controls), we can recover the ATE.

-   Intuition: Reweighting creates a pseudo‑population where everyone looks like a complier in terms of their compliance propensity—effectively undoing the LATE restriction.

-   The estimator requires:
    1.  A model for the compliance score (e.g., probit or logit).
    2.  The exclusion restriction and monotonicity (same as LATE).
    3.  No defiers (monotonicity) and correct specification of the compliance score model.

---
### Implementation in R: `icsw` Package

-   The `icsw` package provides functions for inverse compliance score weighting.

-   Key function: `icsw.tsls()` (two‑stage least squares with inverse compliance score weighting).

-   Steps:
    1.  Estimate compliance scores from a model of treatment on instrument and covariates.
    2.  Construct weights: $w_i = \frac{Z_i}{\hat{\pi}_i} + \frac{1-Z_i}{1-\hat{\pi}_i}$.
    3.  Run weighted IV or weighted OLS (depending on the estimator).

-   Bootstrap is used for standard errors (because compliance scores are estimated).

---
### Application: Peru Emotions Example

```{r peru_icsw_setup, echo = TRUE}
# Reload the trimmed data if needed
peruemotionstrim <- na.omit(data.frame(
  enojado = peruemotions$enojado,
  outsidervote = peruemotions$outsidervote,
  simpletreat = peruemotions$simpletreat,
  Cuzco = peruemotions$Cuzco,
  age = peruemotions$age
))

# Load icsw (install if needed)
# packageurl <- "http://cran.r-project.org/src/contrib/Archive/icsw/icsw_1.0.0.tar.gz"
# install.packages(packageurl, repos=NULL, type="source")
library(icsw)
```

---
```{r peru_icsw_run, echo = TRUE}
# Estimate ATE using inverse compliance score weighting
# D = treatment (enojado), Y = outcome (outsidervote)
# Z = instrument (simpletreat), X = covariates (including intercept)
# W = covariates for compliance score model (can be same as X)

exp.reweight <- with(peruemotionstrim, 
                     icsw.tsls(D = enojado, 
                               Y = outsidervote, 
                               Z = simpletreat, 
                               X = cbind(1, Cuzco),      # Exogenous in outcome model
                               W = cbind(Cuzco, age),    # Covariates for compliance score
                               R = 100))                  # Bootstrap replications
```

---
### Interpreting the Results

```{r peru_icsw_results, echo = TRUE}
# View the ATE estimate and bootstrapped SE
exp.reweight$coefficients
exp.reweight$coefs.se.boot
```

-   The ATE estimate for the effect of anger on outsider voting is `r round(exp.reweight$coefficients, 4)`.

-   Compare with the LATE estimate from IV: `r round(iv_coef, 4)`.

-   The ATE is slightly larger than the LATE. This could mean:
    -   Always‑takers and never‑takers have larger effects than compliers (or opposite signs).
    -   Sampling variability (check confidence intervals).
    -   Model misspecification in the compliance score.

---
### Comparison: LATE vs. ATE

| Estimand | Estimate | Interpretation |
|----------|----------|----------------|
| LATE (IV) | `r round(iv_coef, 4)` | Effect for compliers—those whose anger is activated by the prime. |
| ATE (ICSW) | `r round(exp.reweight$coefficients, 4)` | Effect for the entire population, assuming the compliance score model is correct. |

-   The difference suggests possible treatment effect heterogeneity: compliers may respond differently than always‑takers/never‑takers.

-   But we must interpret with caution: the ATE estimate relies on correct specification of the compliance score model.

---
### Assumptions and Limitations of ICSW

-   **Same as LATE**: exclusion, monotonicity, independence.
-   **Additional assumptions**:
    -   Correctly specified model for the compliance score.
    -   Overlap: compliance scores bounded away from 0 and 1 (no units with $\pi_i = 0$ or $1$).
    -   Consistency of the compliance score estimator.

-   **Practical limitations**:
    -   Can be sensitive to model choice.
    -   Bootstrap standard errors may be unstable with small samples.
    -   Not yet a standard part of the applied toolkit (use with caution and robustness checks).

---
### When Might You Use ICSW?

-   When you have a strong reason to believe that the LATE is not generalizable and you want to estimate the population ATE.

-   As a robustness check to see if the LATE and ATE differ substantially—suggesting treatment effect heterogeneity.

-   In settings with rich covariates that can reliably predict compliance.


---
### Common Mistake 1: Lagged Dependent Variables as Instruments

-   A tempting but usually invalid instrument: using a lag of an endogenous variable as an instrument for itself.

-   Example: In a panel model with $Y_{it} = \beta X_{it} + \gamma Y_{i,t-1} + \epsilon_{it}$, using $X_{i,t-1}$ as an instrument for $X_{it}$.

-   Why this fails:
    -   If $X$ is serially correlated (it usually is), then $X_{i,t-1}$ is correlated with past shocks, which may persist and affect current $Y$.
    -   The exclusion restriction is violated unless the lag is truly exogenous.

-   **When might it work?** In models with strict exogeneity and no serial correlation in errors—rare in practice.

---
### Common Mistake 2: "Many Weak Instruments" Bias

-   Using many weak instruments can actually **increase** bias in 2SLS.

-   With many instruments, the first stage overfits, and 2SLS converges to OLS (which is biased).

-   Rule of thumb: the number of instruments should be small relative to the sample size, and first-stage F should be large.

-   If you have many instruments, consider:
    -   LIML (Limited Information Maximum Likelihood)
    -   Jackknife IV (JIVE)
    -   Weak-instrument robust inference (AR, CLR)

---
### Common Mistake 3: Forgetting to Cluster Standard Errors

-   If the instrument varies at a higher level than the unit of analysis (e.g., state-level policy instrument with individual-level outcomes), standard errors must be clustered at the instrument level.

-   Example: Using state-level minimum wage as an instrument for individual wages—errors are correlated within states.

-   Failure to cluster can lead to severely understated standard errors and false positives.

---
```{r cluster_example, echo = TRUE}
# In ivreg, clustering is not built-in; use lmtest and sandwich
library(lmtest)
library(sandwich)

# Suppose we have clustered data (e.g., by country in MSS example)
# coeftest(migueliv, vcov = vcovCL, cluster = ~country_name)

# Always think about the level at which treatment/instrument varies!
```

---
### Common Mistake 4: Interpreting IV as ATE Without Justification

-   IV estimates LATE, not ATE, unless:
    -   Treatment effects are constant across units.
    -   There are no always‑takers or never‑takers (unlikely).
    -   The instrument affects everyone (i.e., compliance rate = 1).

-   Researchers often slip into language like "the effect of X on Y" without specifying "for compliers."

-   Be precise: "Among those whose treatment status is changed by the instrument, the estimated effect is ..."

---
### Summary: IV Pitfalls to Avoid

| Mistake | Why It's Problematic | Better Approach |
|---------|----------------------|-----------------|
| Lagged DV as instrument | Usually violates exclusion | Use external instruments or GMM |
| Many weak instruments | Increases bias | LIML, JIVE, weak-IV robust tests |
| Ignoring clustering | SEs too small | Cluster at instrument level |
| Interpreting as ATE | Overgeneralization | Be precise: LATE for compliers |

-   IV is powerful but requires careful justification and diagnostics.
-   When in doubt, remember Wright's original insight: find a genuine exogenous shifter.


---
### When Might 2SLS or IV Be a Good Idea?


-   If there's a true randomization or something close in the world that you can take
    advantage of -- but there still might be down-sides.

-   As a Hausmann test/robustness check when OLS may be biased, to see if your most important results hold up under some alternative assumptions.

-   Never without attention to assumptions.

-   If a reviewer demands it of you? (Even then, it can make things worse unless it's intellectually justified.)




