---
title: "Political Science 406 Lab 5: Natural Experiments, Part 2"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
  pdf_document: default
urlcolor: blue
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,                 # don't show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = FALSE,                # set to TRUE to save results from last compilation
  fig.align = "center",         # center figures
  attr.source = '.numberLines'  # add line numbers to code
)

library(tidyverse)       # load libraries you always use here

set.seed(27)             # make random results reproducible
```

# Due Date: May 9, 2025

[Submit your lab here.](https://canvas.northwestern.edu/courses/230286/assignments/1562397)

### Problem 1: Rainfall, Economic Shocks, and Civil Conflict

We are going to replicate an instrumental variables analysis of civil conflict in Africa, using rainfall as
an instrument for economic growth. Consider skimming [the article](https://doi.org/10.1086/421174), by Edward Miguel, Shanker Satyanath, and Ernest Sergenti. Download [replication data, which comes with variable definitions](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27324).

Please replicate model 5 of table 4. Do you get the same coefficient estimates? How about standard
errors? If there is a difference, can you explain why?

How strong is the instrument? Are there any obvious difficulties with the estimate?

Assume that the instrument is unproblematic. What is the causal meaning of the estimate?

### Problem 2: Regression Discontinuity and Partisan Sheriffs

Our [original article](https://doi.org/10.1017/S0003055419000613) here is by Thompson. 

Replicate one of the key RDD analyses from the article as closely as you can, using the [replication data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/CFASH6).

Taking advantage of the things you learned in class about RDD, are there ways you would have changed
or improved the statistical analysis? Make those changes, and report the result, emphasizing any effects this
would have on the substantive conclusions to be drawn.

A good tool for carrying out regression-discontinuity analysis is the rdd library in R, which is documented [here](https://cran.r-project.org/web/packages/rdd/).

A straightforward way to quickly analyze an RDD is like this:

```{r, echo=TRUE, eval=FALSE}
rdoutput <- RDestimate(OUTCOME_VARIABLE ~ ASSIGNMENT_VARIABLE, cutpoint=0.5, data=DATAFRAME)
```

Then, having run that, you can use summary and plot to get good, quick results. There may still be things you want to adjust, but this is the quickest starting point.