---
title: "Political Science 406 Lab 6: Missing Data"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
  pdf_document: default
urlcolor: blue
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,                 # don't show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = FALSE,                # set to TRUE to save results from last compilation
  fig.align = "center",         # center figures
  attr.source = '.numberLines'  # add line numbers to code
)

library(tidyverse)       # load libraries you always use here

set.seed(27)             # make random results reproducible
```

# Due Date: May 10, 2024

[Submit your lab here.](https://canvas.northwestern.edu/courses/211656/assignments/1392658)

### Problem 1: Democracy and Inequality

Return to the inequality data set from Lab 1. Repeat the following analysis using multiple imputation
to handle missing data. How, if at all, do the results differ? What can we conclude?

```{r, echo=TRUE, eval=FALSE}
ineqlm3 <- lm(Gini ~ Polity + log(GDP) + Polity:log(GDP), data=inequality)
```

Try to put together a Tobit-2 selection model instead, positing an explanation for missingness the best
that you can with the available data. Describe your assumptions and explain your results.

### Problem 2: Rainfall, War, and Missing Data

Repeat the rainfall and civil wars analysis from Lab 5, multiply imputing the data. How, if at all, do
your results change using multiple imputation? What do the results mean?
