---
title: "Political Science 406 Lab 7: Synthetic Control"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
  pdf_document: default
urlcolor: blue
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,                 # don't show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = FALSE,                # set to TRUE to save results from last compilation
  fig.align = "center",         # center figures
  attr.source = '.numberLines'  # add line numbers to code
)

library(tidyverse)       # load libraries you always use here

set.seed(27)             # make random results reproducible
```

# Due Date: May 23, 2025

[Submit your lab here.](https://canvas.northwestern.edu/courses/230286/assignments/1562447)

### Problem 1: Trump and Global Warming

Can we use a synthetic controls strategy to come up with a statistical test of the effect of Trump’s
election in 2016 on carbon emissions by the U.S.? Provide a proposed research design complete with
necessary data and estimator, explain the required assumptions, and discuss the strengths and weaknesses
of your proposal. Then show your results if possible.

The library for synthetic controls in R is Synth, documented [here](https://cran.r-project.org/web/packages/Synth/index.html).

This library works in somewhat unusual ways relative to most other stuff in R. Most of the work is done inside the dataprep command:

```{r, echo=TRUE, eval=FALSE}
example.prep <- dataprep(
    foo=DATAFRAME, 
    predictors=c(
      "BACKGROUND_VARIABLE_1",
      "BACKGROUND_VARIABLE_2",
      "BACKGROUND_VARIABLE_3",
      "BACKGROUND_VARIABLE_4"
    ),
    dependent="OUTCOME_VARIABLE",
    unit.variable="SEQUENTIAL_NUMBER_THAT_IDENTIFIES_CASES",
    time.variable="SEQUENTIAL_NUMBER_THAT_IDENTIFIES_TIME_PERIODS", 
    treatment.identifier="NAME_OF_TREATMENT_CASE", 
    controls.identifier=LIST_OF_NAMES_OF_CONTROL_CASES,          
    time.predictors.prior=c(1975:1990), #The first number should be the time 
                          #period that the data starts, and the second
                          #should be the time period before the treatment.
    time.optimize.ssr=c(1975:1989), #Optimize over the period before the treatment.
    time.plot=c(1975:2008), #Plot all the data.
    unit.names.variable="ALPHABETICAL_VARIABLE_THAT_IDENTIFIES_CASES"
)
```

Then, once you've set up the scenario, you can run the synthetic control:

```{r, echo=TRUE, eval=FALSE}
example.synth <- synth(example.prep)
```

Finally, we usually like to get graphs for this process:

```{r, echo=TRUE, eval=FALSE}
path.plot(synth.res=example.synth, dataprep.res=example.prep,
           Ylab="OUTCOME_VARIABLE", Legend=NA, tr.intake=1991,
           Ylim=c(6,8) , Main="NAME_OF_TREATMENT_CASE"
)```

You may need to change the tr.intake to match your treatment year, and you may need to change the Ylim to match the range of your outcome variable.