---
title: "Political Science 406 Lab 8: Difference-in-Differences"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
  pdf_document: default
urlcolor: blue
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,                 # don't show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = FALSE,                # set to TRUE to save results from last compilation
  fig.align = "center",         # center figures
  attr.source = '.numberLines'  # add line numbers to code
)

library(tidyverse)       # load libraries you always use here

set.seed(27)             # make random results reproducible
```

# Due Date: May 30, 2025

[Submit your lab here.](https://canvas.northwestern.edu/courses/230286/assignments/1562450)

### Problem 1: Democracy and Education

In this analysis, I will ask you to use data from the Alvarez et al. data set on regimes and economic
performance worldwide. The data are available as aclppanel.csv, and you can also access [a codebook](http://www.u.arizona.edu/~mishler/PrzeworskiCodebook.PDF).

Using these data, try to determine whether dictatorship/democracy (REG) has a positive or negative
causal effect on average education levels (EDT). Structure the analysis as a difference-in-differences setup,
and deal with any problems that arise. Describe your findings.

The main contemporary R library we used that can accommodate treatments that happen at different points in time is fect, documented [here](https://cran.r-project.org/web/packages/fect/index.html). The command can be run as follows:

```{r, echo=TRUE, eval=FALSE}
output.ect <- fect(OUTCOME.VARIABLE ~ TREATMENT.VARIABLE, index = c("COUNTRY", "YEAR"), force = "two-way",
               data = DATAFRAME, method="fe",min.T0=1, se=1)
```

It can be interesting to try different methods. gsynth instead of fe uses synthetic controls to estimate the diff-in-diff, for example, but only works when cases don't switch back to control after becoming treated. mc uses a matrix completion method, and polynomial allows nonlinear time trends.